---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300,
  fig.width = 8,
  fig.height = 5
)
```

# move2imu

<!-- badges: start -->
[![R-CMD-check](https://github.com/move2universe/move2imu/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/move2universe/move2imu/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/move2universe/move2imu/graph/badge.svg)](https://app.codecov.io/gh/move2universe/move2imu)
[![CRAN status](https://www.r-pkg.org/badges/version/move2imu)](https://CRAN.R-project.org/package=move2imu)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

move2imu aims to standardize the storage and analysis of biologging 
inertial measurement unit (IMU) data, including accelerometer, magnetometer,
and gyroscope records. The package integrates with [move2](https://bartk.gitlab.io/move2/), enabling
standardized data processing workflows and allowing IMU data to be analyzed alongside
other observations, including location records.

## Installation

move2imu does not yet exist on CRAN. Instead, you can install the development
version directly:

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("move2universe/move2imu")

# Or, if remotes is already installed:
remotes::install_github("move2universe/move2imu")
```

## Usage

Extract and standardize IMU bursts from a `move2` object or a `data.frame`:

```{r}
library(move2imu)
library(move2)

# Extract acceleration data from gulls data
a <- as_acc(gulls())
a <- a[!is.na(a)]

head(a)

# Overview of acceleration bursts
summary(a)
```

Calibrate tags:

```{r}
# Standardize raw ADC counts to physical units with a built-in tag calibration
a <- transform_imu(a, acc_calibration("ornitela", units = "standard_free_fall"))

head(a)
```

Visualize sampling regimes:

```{r sampling-effort, fig.alt = "Sampling effort for nine albatross tracks over a two-minute window. Each track is drawn as its own row, with shaded bins marking the times at which acceleration samples were recorded."}
# Visualize sampling patterns in your data
alb <- albatrosses()

plot_sampling_effort(
  acc = as_acc(alb),
  ids = mt_track_id(alb),
  from = as.POSIXct("2008-07-27 00:00:00", tz = "UTC"),
  to = as.POSIXct("2008-07-27 00:02:00", tz = "UTC")
)
```

## Getting help + Contributing

We welcome feedback and contributions. If you encounter a bug or have 
specific feature requests, please create an issue on [GitHub](https://github.com/move2universe/move2imu).
