---
title: "Introduction to move2imu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to move2imu}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  out.width = "80%"
)

local({
  hook_output <- knitr::knit_hooks$get("output")
  knitr::knit_hooks$set(output = function(x, options) {
    n <- options$out.lines
    if (!is.null(n)) {
      lines <- strsplit(x, "\n", fixed = TRUE)[[1]]
      if (length(lines) > n) {
        x <- paste(c(lines[seq_len(n)], "#> ..."), collapse = "\n")
      }
    }
    hook_output(x, options)
  })
})
```

```{css, echo = FALSE}
.html-widget {
  margin-left: auto !important;
  margin-right: auto !important;
  width: 80% !important;
}
```

Biologging devices often record information about an animal's behavior using
an inertial measurement unit (IMU) that collects acceleration, magnetometer,
and/or orientation data. Because collecting these data can be battery- and storage-intensive, they are often recorded intermittently in rapid sets of samples at a fixed frequency, which we will refer to as *bursts*. These bursts of IMU
measurements provide regular snapshots of an animal's behavior and activity.

(Note that while some IMUs allow you to record data continuously 
for longer periods of time, we still use the term *burst* to refer to a
distinct set of fixed-frequency IMU data samples, regardless of how long
such a "burst" lasts.)

move2imu provides an R vector representation of these bursts so that they
can be easily manipulated, analyzed, and linked to other tracking data---for
instance, those stored in a `move2` object. The package currently supports three sensor
types:

- `acc()` for accelerometer bursts
- `mag()` for magnetometer bursts
- `gyro()` for gyroscope bursts

Each of these shares the same underlying structure. For the rest of this
vignette, we will use acceleration data as an example.

```{r load, message=FALSE}
library(move2imu)
library(move2)
library(dplyr)
```

## Burst structure

You can create an `acc` vector from scratch by providing a list of burst
matrices (containing the raw acceleration data) and the associated sampling
frequency for those observations. Each burst can contain up to three recording
axes: `X`, `Y`, and/or `Z`. Note that the frequency must be provided in frequency units (in this case, `"Hz"`). move2imu uses the [units](https://r-quantities.github.io/units/) package to support unit manipulation.

```{r create}
raw_acc <- cbind(
  X = sin(1:30 / 10),
  Y = cos(1:30 / 10),
  Z = 1:30 / 10
)

a1 <- acc(
  bursts = list(raw_acc),
  frequency = units::as_units(40, "Hz")
)

a1
```

The printed output of the vector shows the number of bursts as well as the
mean value per axis for each burst. Internally, the raw data matrix is
preserved.

Bursts can also be associated with a timestamp marking the start time of the
burst:

```{r}
a2 <- acc(
  bursts = list(
    cbind(X = sin(1:20 / 10), Y = cos(1:20 / 10)),
    cbind(X = sin(1:20 / 10 + 2), Y = cos(1:20 / 10 + 3))
  ),
  frequency = units::as_units(30, "Hz"),
  start = as.POSIXct(c("2026-06-01 00:00:00", "2026-06-01 00:01:00"), tz = "UTC")
)
```

Multiple `acc` vectors can be combined with `c()`, even if their sampling
frequency, axes, or other properties vary:

```{r combine}
a <- c(a1, a2)

a
```

## Loading IMU data

Of course, in most cases you won't be building bursts from scratch, but will
load these data from external sources. move2imu has been designed to work in
tandem with [move2](https://bartk.gitlab.io/move2/) and 
[Movebank](https://www.movebank.org) and will automatically parse IMU data 
stored in one of Movebank's standard storage formats 
(see `movebank_acc_colsets()`).

For example, we'll load a sample dataset of albatross tracks stored as a `move2`:

```{r}
# Load sample Movebank data into a `move2`
alb <- albatrosses()

alb
```

Then, to extract the acceleration data from this object, we simply call `as_acc()`:

```{r}
a <- as_acc(alb)

a
```

`as_acc()` automatically identifies the relevant acceleration data in the
albatross data and converts them into bursts with associated frequencies and
start timestamps.

By default, rows in the `move2` with no acceleration data receive 
missing (`NA`) `acc` bursts. This makes it easy to keep acceleration 
bursts linked to other metadata stored in the `move2`, like GPS location:

```{r}
alb |>
  mutate(acceleration = a) |>
  select(acceleration)
```

### Input data formats

The Movebank data model supports several storage formats for acceleration
data. However, move2imu handles this heterogeneity out of the box:

```{r, out.lines = 10}
# Sample Movebank data
gul <- gulls()

as_acc(gul)
```

Even though the albatross and gulls data sources store their acceleration data
differently, because each adheres to the Movebank data model, `as_acc()`
seamlessly parses the data into a set of `acc` burst matrices.

For more details about the ways Movebank stores acceleration data, see 
`movebank_acc_colsets()`.

### Fine-tuning data import

If you have your own data stored using column names that don't adhere
to the Movebank data model, you can still use `as_acc()` to load them, provided
that the data themselves are stored in a recognizable format (see 
`imu_colset()`). Use `imu_colset()` to define the column names that correspond to each
axis of acceleration data in your input data source:

```{r, eval = FALSE}
# Extract acc data from the "acc_x" and "acc_y" columns
a <- as_acc(x, colset = imu_colset(x = "acc_x", y = "acc_y"))
```

## Exploring bursts

For a quick summary of an IMU vector, use `summary()`. This shows the
range of several key burst properties as well as the quantiles of the
inter-burst time intervals and the actual recorded burst values.

```{r}
a <- as_acc(albatrosses())

summary(a)
```

For a more detailed look at these properties, you can use move2imu's
dedicated accessors. For instance:

```{r}
# Number of samples per burst
n_samples(a)

# Burst duration, in seconds
burst_dur(a)

# Check whether all bursts have same frequency, samples, axes, units
is_uniform(a)

# and more...
```

If you need to operate on the burst data or metadata directly, they can easily be accessed as well:

```{r, out.lines = 10}
bursts(a)

freqs(a)

starts(a)
```

These can be used to easily explore the distributions of these properties:

```{r summary_hist, fig.width=6, fig.height=4}
# Distribution of inter-burst intervals
hist(burst_intervals(a), main = "Burst intervals (s)")

# All values
hist(unlist(bursts(a)), main = "Sample values")
```

### Plotting sampling effort

Many tags change their sampling settings over the course of a study. To
visualize how the IMU data were sampled, use `plot_sampling_effort()`. This can
reveal periods where different sensors or tracks collected data simultaneously,
where tags stopped and started collecting data, and how densely certain
time ranges were sampled.

```{r}
track_ids <- mt_track_id(albatrosses())

plot_sampling_effort(acc = a, ids = track_ids)
```

Modify the width of the aggregation bins and the plot's time range to explore
time periods in more detail:

```{r}
start <- as.POSIXct("2008-07-27 00:00:00", tz = "UTC")
end <- as.POSIXct("2008-07-27 00:02:00", tz = "UTC")
width <- units::set_units(0.3, "s")

plot_sampling_effort(
  acc = a,
  ids = track_ids,
  bin_width = width,
  from = start,
  to = end
)
```

`plot_sampling_effort()` accepts both `imu` objects as well as timestamp
vectors, so you can also plot other sensors based on their recorded timestamps
(e.g. GPS coordinates):

```{r, fig.height=6}
# Mask out timestamps to get only those where a GPS location was recorded
gps <- replace(mt_time(albatrosses()), sf::st_is_empty(albatrosses()), NA)

plot_sampling_effort(
  acc = a,
  gps = gps,
  ids = track_ids,
  bin_width = width,
  from = start,
  to = end
)
```

### Plotting IMU traces

You can also plot the values recorded by an IMU vector over time with
`plot_imu_trace()`. This produces an interactive
[`dygraph`](https://rstudio.github.io/dygraphs/) that you can zoom
into with the mouse:

```{r plot, fig.width=6}
a <- as_acc(gulls())

plot_imu_trace(a)
```

You can also pass a single burst (or a subset) for a focused view —
here, the wing beats are clearly visible on the Z axis:

```{r plot_one, fig.width=6}
plot_imu_trace(a[422])
```

The plot is returned unstyled. Additional arguments are passed to
`dygraphs::dygraph()` and additional `dygraphs` layers can be added to
the output plot. For instance, to add axis labels and ensure times are labeled
in UTC instead of the browser's local time zone:

```{r plot_styled, fig.width=6}
plot_imu_trace(a[422], main = "Wing beats", ylab = "Acceleration (g)") |>
  dygraphs::dyOptions(labelsUTC = TRUE)
```

<!-- ## A single burst in 3D -->

<!-- For a closer look at one burst, the matrix of samples is just an -->
<!-- ordinary numeric matrix and can be plotted with base graphics: -->

<!-- ```{r plot3d, fig.width=6, fig.height=6} -->
<!-- b <- bursts(a)[[21]] -->
<!-- e  <- 0.1 -->

<!-- p <- persp( -->
<!--   z = matrix(min(b[, "Z"]) - e, ncol = 2, nrow = 2), -->
<!--   x = range(b[, "X"]), -->
<!--   y = range(b[, "Y"]), -->
<!--   xlim = range(b[, "X"]) + c(-e, e), -->
<!--   ylim = range(b[, "Y"]) + c(-e, e), -->
<!--   zlim = range(b[, "Z"]) + c(-e, e), -->
<!--   border = NA, -->
<!--   xlab = "X", -->
<!--   ylab = "Y", -->
<!--   zlab = "Z", -->
<!--   ticktype = "detailed", cex.axis = 0.65, -->
<!--   theta = -160, phi = 20, expand = 0.5, col = "white", scale = FALSE -->
<!-- ) -->

<!-- lines(trans3d(b[, "X"], b[, "Y"], min(b[, "Z"]) - e, p), col = "tomato") -->

<!-- apply(b, 1, function(x, bottom) { -->
<!--   lines(trans3d(x["X"], x["Y"], c(x["Z"], bottom - e), p), col = "gray") -->
<!-- }, bottom = min(b[, "Z"])) -->

<!-- lines(trans3d(b[, "X"], b[, "Y"], b[, "Z"], p), col = "red") -->

<!-- points(trans3d(b[, "X"], b[, "Y"], b[, "Z"], p), col = "red", pch = 19) -->
<!-- ``` -->

## Next steps

If your data are stored as raw ADC values rather than physical units,
see the [calibration vignette](calibration.html) to learn how to convert them.

See the [programming with bursts](programming.html) vignette for a demonstration
of how to compute values from `acc` bursts and link them back to GPS
coordinates in a `move2` object.
